cardinalR: Generating interesting high-dimensional data structures

### 2025-06-24 paper-cardinalR.Rmd
Updated Gaussian and Linear sections

The `gen_curvybranches(n, p, k)` function generates a dataset of $n$ points forming $k$ curvilinear branches embedded in $p\text{-}D$. The underlying geometry lies in the first two dimensions, while the remaining $(p-2)\text{-}D$ dimensions contain Gaussian noise.

Branch 1 is generated from $X_1 \sim U(0, 1)$, and $X_2 = 0.1 \cdot X_1 + 1 \cdot X_1^2 + \epsilon$, where $\epsilon \sim U(0, 0.05)$. This produces a gently upward-curving parabola in the right half-plane.

Branch 2 is generated from $X_{1} \sim U(-1, 0)$, and $X_2 = 0.1 \cdot X_{1} - 2 \cdot X_{1}^2 + \epsilon$, where $\epsilon \sim U(0, 0.05)$. This creates a steeper, leftward-facing curve in the left half-plane.

Branches $3$ to $k$ are added iteratively. Each new branch begins at a randomly chosen point within a restricted horizontal band: $X_{1} \in [-0.15, 0.15]$ (to ensure connectivity with earlier branches), spans a unit-length interval on $X_{1}$:$X_1 \in [x_{\text{start}}, x_{\text{start}} + 1]$, and has the structure:

@@

For $p > 2$, the remaining dimensions are pure Gaussian noise:

$$
X_j \sim N(0, 0.05^2), \quad \text{for } j = 3, \dots, p.
$$

@@

The `gen_gaussian(n, p, s)` function generates a multivariate Gaussian cloud in $p$-dimensional space, centered at the origin with user-defined covariance structure.

Each point is independently drawn using the multivariate normal distribution with:

$$
X_i \sim N_p(\boldsymbol{0}, \Sigma)
$$

where $\Sigma$ is a user-defined $p \times p$ positive-definite matrix.

@@

The `gen_longlinear(n, p)` function simulates a high-dimensional linear structure with a dominant linear trend and small additive noise.

Each variable $X_i$ is created as:

$$
X_i = \text{scale}_i \cdot \left(0, 1, \dots, n{-}1 + \epsilon\right) + \text{shift}_i
$$

where:

$\text{scale}_i \sim U(-10, 10)$: randomly chosen direction and stretch.

$\text{shift}_i \sim U(-300, 300)$: large offset to spread features apart.

$\epsilon \sim N(0, 0.03 \cdot n)$: Gaussian noise scaled to dataset size to introduce mild irregularity.

>>>>

The `gen_curvybranches(n, p, k)` function generates a dataset of $n$ points forming $k$ curvilinear branches embedded in $p\text{-}D$. The underlying geometry lies in the first two dimensions, while the remaining $(p-2)$ dimensions contain Gaussian noise.

Branch $1$ is generated from $X_1 \sim U(0, 1)$, and $X_2 = 0.1 \cdot X_1 + 1 \cdot X_1^2 + \epsilon$, where $\epsilon \sim U(0, 0.05)$. This produces a gently upward-curving parabola in the right half-plane. Branch $2$ is generated from $X_{1} \sim U(-1, 0)$, and $X_2 = 0.1 \cdot X_{1} - 2 \cdot X_{1}^2 + \epsilon$, where $\epsilon \sim U(0, 0.05)$. This creates a steeper, leftward-facing curve in the left half-plane. Branches $3$ to $k$ are added iteratively. Each new branch $i$ begins at a randomly chosen point within a restricted horizontal band: $X_{1} \in [-0.15, 0.15]$ (to ensure connectivity with earlier branches), spans a unit-length interval on $X_{1}$:$X_1 \in [x_{\text{start}}, x_{\text{start}} + 1]$, and has the structure:

@@

For $p > 2$, Gaussian noise $X_j \sim N(0, 0.05^2)$ is added to embed the $2\text{-}D$ branches into $p\text{-}D$, where $j = 3, \dots, p$.

@@

The `gen_gaussian(n, p, s)` function generates a multivariate Gaussian cloud in $p\text{-}D$, centered at the origin with user-defined covariance structure. Each point is independently drawn using the multivariate normal distribution with $X_i \sim N_p(\boldsymbol{0}, \Sigma)$, where $\Sigma$ is a user-defined $p \times p$ positive-definite matrix.

@@

The `gen_longlinear(n, p)` function simulates a high-dimensional linear structure with a dominant linear trend and small additive noise. Each variable $X_i$ is created as $X_i = \text{scale}_i \cdot \left(0, 1, \dots, n{-}1 + \epsilon\right) + \text{shift}_i$, where $\text{scale}_i \sim U(-10, 10)$ is randomly chosen direction and stretch, $\text{shift}_i \sim U(-300, 300)$ is large offset to spread features apart, and $\epsilon \sim N(0, 0.03 \cdot n)$ is Gaussian noise.

### 2025-06-24 paper-cardinalR.tex
Updated Gaussian and Linear sections

The \texttt{gen\_curvybranches(n,\ p,\ k)} function generates a dataset of \(n\) points forming \(k\) curvilinear branches embedded in \(p\text{-}D\). The underlying geometry lies in the first two dimensions, while the remaining \((p-2)\text{-}D\) dimensions contain Gaussian noise.

Branch 1 is generated from \(X_1 \sim U(0, 1)\), and \(X_2 = 0.1 \cdot X_1 + 1 \cdot X_1^2 + \epsilon\), where \(\epsilon \sim U(0, 0.05)\). This produces a gently upward-curving parabola in the right half-plane.

Branch 2 is generated from \(X_{1} \sim U(-1, 0)\), and \(X_2 = 0.1 \cdot X_{1} - 2 \cdot X_{1}^2 + \epsilon\), where \(\epsilon \sim U(0, 0.05)\). This creates a steeper, leftward-facing curve in the left half-plane.

Branches \(3\) to \(k\) are added iteratively. Each new branch begins at a randomly chosen point within a restricted horizontal band: \(X_{1} \in [-0.15, 0.15]\) (to ensure connectivity with earlier branches), spans a unit-length interval on \(X_{1}\):\(X_1 \in [x_{\text{start}}, x_{\text{start}} + 1]\), and has the structure:

@@

For \(p > 2\), the remaining dimensions are pure Gaussian noise:

\[
X_j \sim N(0, 0.05^2), \quad \text{for } j = 3, \dots, p.
\]

@@

The \texttt{gen\_gaussian(n,\ p,\ s)} function generates a multivariate Gaussian cloud in \(p\)-dimensional space, centered at the origin with user-defined covariance structure.

Each point is independently drawn using the multivariate normal distribution with:

\[
X_i \sim N_p(\boldsymbol{0}, \Sigma)
\]

where \(\Sigma\) is a user-defined \(p \times p\) positive-definite matrix.

@@

The \texttt{gen\_longlinear(n,\ p)} function simulates a high-dimensional linear structure with a dominant linear trend and small additive noise.

Each variable \(X_i\) is created as:

\[
X_i = \text{scale}_i \cdot \left(0, 1, \dots, n{-}1 + \epsilon\right) + \text{shift}_i
\]

where:

\(\text{scale}_i \sim U(-10, 10)\): randomly chosen direction and stretch.

\(\text{shift}_i \sim U(-300, 300)\): large offset to spread features apart.

\(\epsilon \sim N(0, 0.03 \cdot n)\): Gaussian noise scaled to dataset size to introduce mild irregularity.

>>>>

The \texttt{gen\_curvybranches(n,\ p,\ k)} function generates a dataset of \(n\) points forming \(k\) curvilinear branches embedded in \(p\text{-}D\). The underlying geometry lies in the first two dimensions, while the remaining \((p-2)\) dimensions contain Gaussian noise.

Branch \(1\) is generated from \(X_1 \sim U(0, 1)\), and \(X_2 = 0.1 \cdot X_1 + 1 \cdot X_1^2 + \epsilon\), where \(\epsilon \sim U(0, 0.05)\). This produces a gently upward-curving parabola in the right half-plane. Branch \(2\) is generated from \(X_{1} \sim U(-1, 0)\), and \(X_2 = 0.1 \cdot X_{1} - 2 \cdot X_{1}^2 + \epsilon\), where \(\epsilon \sim U(0, 0.05)\). This creates a steeper, leftward-facing curve in the left half-plane. Branches \(3\) to \(k\) are added iteratively. Each new branch \(i\) begins at a randomly chosen point within a restricted horizontal band: \(X_{1} \in [-0.15, 0.15]\) (to ensure connectivity with earlier branches), spans a unit-length interval on \(X_{1}\):\(X_1 \in [x_{\text{start}}, x_{\text{start}} + 1]\), and has the structure:

@@

For \(p > 2\), Gaussian noise \(X_j \sim N(0, 0.05^2)\) is added to embed the \(2\text{-}D\) branches into \(p\text{-}D\), where \(j = 3, \dots, p\).

@@

The \texttt{gen\_gaussian(n,\ p,\ s)} function generates a multivariate Gaussian cloud in \(p\text{-}D\), centered at the origin with user-defined covariance structure. Each point is independently drawn using the multivariate normal distribution with \(X_i \sim N_p(\boldsymbol{0}, \Sigma)\), where \(\Sigma\) is a user-defined \(p \times p\) positive-definite matrix.

@@

The \texttt{gen\_longlinear(n,\ p)} function simulates a high-dimensional linear structure with a dominant linear trend and small additive noise. Each variable \(X_i\) is created as \(X_i = \text{scale}_i \cdot \left(0, 1, \dots, n{-}1 + \epsilon\right) + \text{shift}_i\), where \(\text{scale}_i \sim U(-10, 10)\) is randomly chosen direction and stretch, \(\text{shift}_i \sim U(-300, 300)\) is large offset to spread features apart, and \(\epsilon \sim N(0, 0.03 \cdot n)\) is Gaussian noise.
